---
Author: Bailee Dastugue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

__Goals of this project:__

From a computational biology approach, I wanted to integrate Python functionality within R while satisfying the "please use Python in some capacity" requirement for this project. I used Python to count the number of times a word in the song "Shabop Shalom" was said. I used a connotative lexicon for over 90k words found online (and will eventually be linked *here*) and tidied that data set using R. I joined the two using R and ran some summary statistics. Will soon add some charts that potentially only I will enjoy.

From a more personal perspective, my husband and I share a lot of music on a many collaborative playlists. We started doing this as friends many moons ago and it's been a really wonderful way to document our relationship over the years.

However! We tend to disagree on the *tone* of a song frequently. For example, I find Japanese Breakfast's ["Maybe You're the Reason"](https://open.spotify.com/track/7xKjdvYaJ1tvwNXRTAZs78?si=rSNHKuD_TTqDYeQ8SoQSow) to be a song about finding meaning in love and/or passsion whilst in a strong bout of apathy and existentialism (Aside, is there a  way to run a spell check on an R file? I am in desperate need). As I got a major hip surgery this summer, this song felt Significant and Meaningful as I navigated a lot of the more frustrating aspects of recovery. I added it as a sweet subtweet! So sweet! He thought the vibe was less positive.

I focus much more extensively on lyrics while he is more attuned to the actual music (chords, key, etc). True to form, this project is an attempt to quantify how positive or negative the lyrics of a song are. With a score of 25, "Shabop" seems positive, but it's a relative scoring system entirely dependent on the odd lexicon found online. Would be good to have other songs to measure against - note to self: find the happiest and most depressive songs to score at a later date.


```{R}
library(readr)
lexi <- read_csv("~/desktop/website/content/blog/lexi.csv")
library(reticulate)
library(dplyr)
library(tidyr)
library(tidyverse)
lexi
lexi1 <- lexi %>% rename(word = Word) %>% rename(conn = connotation) %>% glimpse()
lexi2 <- lexi1 %>% separate(word, into = c("word", "type"), sep = "_") %>% mutate_if(is.character, as.factor) %>% glimpse()
```

```{r}
shabop <- paste(read_lines("shabop.txt"), collapse = " ")
shabop
```

```{python}
import re
shabop2 = re.sub(r"[.—,;'\")()?-]","",r.shabop)
print(shabop2)
```

```{python}
shabop3 = re.split(r" +", shabop2)
lower_list=[] #empty list
for word in shabop3:
  temp = word.lower()
  lower_list.append(temp)
print(lower_list)
```

```{python}
counts = {}
for c in lower_list:
    if c in counts:
        counts[c]+=1
    else:
        counts[c]=1
for c in counts:
   print (c, 'appears', counts[c],"times")
```


```{python}
list = []
for c in counts:
   new = (c, counts[c])
   list.append(new)
```


```{r}
times <- py$list %>% as.data.frame() %>% select_if(is.integer) %>% pivot_longer(contains(c("X")), names_to = "name", values_to = "times") %>% mutate(id = row_number()) %>% select(-name)
times
words <- py$list %>% as.data.frame() %>% select_if(is.factor) %>% pivot_longer(contains(c("X")), names_to = "name", values_to = "word") %>% mutate(id = row_number()) %>% select(-name)
words
full <- words %>% left_join(times) %>% select(-id)
```

```{r}
full %>% left_join(lexi2)
```
After joining the song lyrics with the connotative lexicon, the number of observations actually increased - which actually surprised me. This is because the lexicon accounts for the classification of word (i.e. pronoun, verb, adjective). Part of me wants to do away with that system entirely, but each word can have different connotations for each classification. To neglect those different connotations just because I am lazy, wouldn't accurately represent the actual song. So, persist we must. To essentially side step this issue, I created a numeric column to correspond to each connotation: +1 for a positive connotation, -1 for negative, and 0 for a neutral connotation. By taking the average of these values for each word, I get a more comprehensive connotation score without over-representing words that have multiple classifications.

```{r}
lexi3 <- lexi2 %>% mutate(value = ifelse(conn == "positive", 1, ifelse(conn == "negative", -1, 0))) %>% select(-conn) %>% group_by(word) %>%  summarise_at(vars(-type), funs(mean(., na.rm=TRUE))) %>% mutate_if(is.integer, as.double)
lexi3 
shalom <- full %>% left_join(lexi3)
shalom
shalom1 <- shalom %>% mutate(weighted_value = times*value) %>% na.omit
```
Main short coming of this lexicon: it doesn't have every word used in the song, particularly overlooking the plural form of each word. Perhaps I will go back and see if I get more comprehensive results if I coerce each word into its singular form, but I think that could get tricky. I could also add an 's' to every word in the lexicon -- which might be less tricky, even if it comes up with nonsense frequently.

```{r}
shalom1 %>% select(weighted_value) %>% summarise_all(sum)
```

Overall, this song is positive - thank god too like how was I going to tell my husband that this song - the first song that either of us ever added to one of our shared playlists - is actually a bummer ???? Thank you R

